### 2025-03-04 sec/4_experimentsandresults.tex
add diffusion model architecture

\section{Experiments and Results}
\label{sec:experiments-and-results}

\subsection{Expert Survey: Indistinguishability of Synthetic Images}
\label{subsec:expert-survey:-indistinguishability-of-synthetic-images}
The expert survey results revealed a striking inability of microscopy experts to consistently distinguish synthetic brightfield microscopy images generated by our diffusion model from real images.
The overall classification accuracy all 11 participants and 30 images was remarkably close to chance level, reaching only 50\%.
This near-random performance strongly indicates the high degree of realism achieved by our synthetic image generation approach.
Figure~\ref{fig:conf_matrix} presents the normalized confusion matrix, visually demonstrating the balanced distribution of correct and incorrect classifications for both real and generated images.
Individual image accuracies varied significantly, ranging from approximately 18\% to 90\%, highlighting the inherent variability within both real and synthetic image sets and the complexity of the discrimination task.
Analysis of expert explanations revealed that experts considered subtle image features related to cell appearance, background texture, edge clarity, and perceived noise or artifacts when attempting to differentiate the images.
Prominent terms like ``cell,'' ``background,'' and ``edge'' indicate a focus on core microscopy image components, while terms such as ``pattern,'' ``perfect,'' ``artifacts,'' ``blurry,'' and ``contrast'' suggested that experts were searching for subtle imperfections or stylistic inconsistencies that might betray a synthetic origin.
However, these cues proved unreliable, ultimately leading to the near-chance level classification accuracy and underscoring the perceptual realism of our diffusion-generated brightfield microscopy images.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.75\linewidth]{images/norm_confusion_matrix}
    \caption{Normalized confusion matrix of expert survey classifications. Overall accuracy is approximately 50\%, demonstrating the significant difficulty for experts in distinguishing real and synthetic brightfield microscopy images.}
    \label{fig:conf_matrix}
\end{figure}

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease emerges for models trained with increasing proportions of replaced synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.

Continuing the analysis of object detection performance, we turn on the datasets where synthetic images were \textbf{added} to the real training data (\textit{scc\_add\_10}, \textit{scc\_add\_30}, \textit{scc\_add\_50}).
Examining mAP\@50 in Table~\ref{tab:model-performances}, we observe that models trained on these augmented datasets generally maintain a high level of performance, comparable to the baseline performance achieved with the \textit{scc\_real} dataset as well.
For instance, the YOLOv8s model shows a consistent mAP@50 around 0.8949-0.8952 across \textit{scc\_add\_10}, \textit{scc\_add\_30}, and \textit{scc\_add\_50}, similar to the 0.8947 of \textit{scc\_real}.
Notably, the YOLOv8m model achieves a mAP@50 of 0.9034 on \textit{scc\_add\_30}, closely matching the 0.9038 of the \textit{scc\_real} baseline.
The larger models, YOLOv8x, YOLOv9c, and YOLOv9e, also exhibit this trend of maintained high mAP@50 when trained with added synthetic data, suggesting that supplementing real data with synthetic images does not degrade the detection accuracy at this IoU threshold.

>>>>

\section{Experiments and Results}
\label{sec:experiments-and-results}
After thorough evaluation of the various unconditional diffusion model architectures, noise schedulers, and hyperparameters,
a final configuration was selected for the image generation task.
The chosen model is the \textit{scc\_small} architecture, which demonstrated a good balance between performance and computational efficiency.
It consists of 5 block out channels (128, 128, 256, 256, 512) and 5 block in channels (128, 128, 256, 256, 512), with no attention layers and a total of 70.1M trainable parameters.
It was trained for 350 epochs in two days and 7 hours, with the above mentioned hyperparameters and noise scheduler settings.
This model, despite its relatively smaller size, achieved competitive results in terms of image quality and fidelity.

\subsection{Expert Survey: Indistinguishability of Synthetic Images}
\label{subsec:expert-survey:-indistinguishability-of-synthetic-images}
The expert survey results revealed a striking inability of microscopy experts to consistently distinguish synthetic brightfield microscopy images generated by our diffusion model from real images.
The overall classification accuracy of all 11 participants and 30 images was remarkably close to chance level, reaching only 50\%.
This near-random performance strongly indicates the high degree of realism achieved by our synthetic image generation approach.
Figure~\ref{fig:conf_matrix} presents the normalized confusion matrix, visually demonstrating the balanced distribution of correct and incorrect classifications for both real and generated images.
Individual image accuracies varied significantly, ranging from approximately 18\% to 90\%, highlighting the inherent variability within both real and synthetic image sets and the complexity of the discrimination task.
Analysis of expert explanations revealed that experts considered subtle image features related to cell appearance, background texture, edge clarity, and perceived noise or artifacts when attempting to differentiate the images.
Prominent terms like ``cell,'' ``background,'' and ``edge'' indicate a focus on core microscopy image components, while terms such as ``pattern,'' ``perfect,'' ``artifacts,'' ``blurry,'' and ``contrast'' suggested that experts were searching for subtle imperfections or stylistic inconsistencies that might betray a synthetic origin.
However, these cues proved unreliable, ultimately leading to the near-chance level classification accuracy and underscoring the perceptual realism of our diffusion-generated brightfield microscopy images.
\begin{figure}
    \centering
    \includegraphics[width=0.75\linewidth]{images/norm_confusion_matrix}
    \caption{Normalized confusion matrix of expert survey classifications. Overall accuracy is approximately 50\%, demonstrating the significant difficulty for experts in distinguishing real and synthetic brightfield microscopy images.}
    \label{fig:conf_matrix}
\end{figure}

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease emerges for models trained with increasing proportions of replaced synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.
\begin{figure*}
    \centering
    % Real images group with spanning subcaption
    \begin{minipage}[b]{0.49\textwidth}
        \centering
        \captionsetup{type=figure}
        \subcaptionbox{Sample Real Images\label{fig:real}}{%
            \begin{minipage}[t]{0.45\linewidth}
                \includegraphics[width=\linewidth]{images/real_cell_image_1561}
                \label{fig:real1}
            \end{minipage}\hfill
            \begin{minipage}[t]{0.45\linewidth}
                \includegraphics[width=\linewidth]{images/real_cell_image_0987}
                \label{fig:real2}
            \end{minipage}
        }
    \end{minipage}\hfill
    % Generated images group with spanning subcaption
    \begin{minipage}[b]{0.49\textwidth}
        \centering
        \captionsetup{type=figure}
        \subcaptionbox{Sample Generated Images\label{fig:gen}}{%
            \begin{minipage}[t]{0.45\linewidth}
                \includegraphics[width=\linewidth]{images/sample_00885_37_png.rf.e2420f6e7550dc08ab1fc2688adc84f7}
                \label{fig:gen1}
            \end{minipage}\hfill
            \begin{minipage}[t]{0.45\linewidth}
                \includegraphics[width=\linewidth]{images/sample_00870_40_png.rf.7c55b5f72263d273e38102bf8a0db76a}
                \label{fig:gen2}
            \end{minipage}
        }
    \end{minipage}
    \caption{Visual comparison between real-world samples \subref{fig:real} and generated samples \subref{fig:gen}}
    \label{fig:comparison}
\end{figure*}
Continuing the analysis of object detection performance, we turn on the datasets where synthetic images were \textbf{added} to the real training data (\textit{scc\_add\_10}, \textit{scc\_add\_30}, \textit{scc\_add\_50}).
Examining mAP\@50 in Table~\ref{tab:model-performances}, we observe that models trained on these augmented datasets generally maintain a high level of performance, comparable to the baseline performance achieved with the \textit{scc\_real} dataset as well.
For instance, the YOLOv8s model shows a consistent mAP@50 around 0.8949-0.8952 across \textit{scc\_add\_10}, \textit{scc\_add\_30}, and \textit{scc\_add\_50}, similar to the 0.8947 of \textit{scc\_real}.
Notably, the YOLOv8m model achieves a mAP@50 of 0.9034 on \textit{scc\_add\_30}, closely matching the 0.9038 of the \textit{scc\_real} baseline.
The larger models, YOLOv8x, YOLOv9c, and YOLOv9e, also exhibit this trend of maintained high mAP@50 when trained with added synthetic data, suggesting that supplementing real data with synthetic images does not degrade the detection accuracy at this IoU threshold.
